# README

Workflow to upload DesFert stem-length data to the urbancndep database.

```{r}
#| eval: true

source("helper_load_table.R")
source("helper_identify_new_missing.R")
source("helper_annotate_new_missing.R")
source("helper_safe_transaction.R")

```


# temporary schema -------------------------------------------------------------

```{r}
#| eval: true

DBI::dbExecute(pg, "DROP SCHEMA IF EXISTS stems_temp CASCADE ;")
DBI::dbExecute(pg, "CREATE SCHEMA IF NOT EXISTS stems_temp ;")

```


# add raw data to temp postgres tables -----------------------------------------

```{r}
#| eval: true

data_tables <- c("new", "old", "plots_plants", "plot_notes", "shrub_dimensions")

purrr::walk(
  .x = data_tables,
  .f = ~ helper_load_table(table_name = .x)
)
```


# data manipulation (SQL) -----------------------------------------------------

run SQL run sequentially!

alter stems_temp.plots_plants as appropriate
- change survey_date to a date (may not be needed)
- add shrub ID from urbancndep
- prep for and add shrub_ids to the plots_plants tablet-derived data
- prep for and add stem_id generated from insert of new stems to plots_plants (a later step)

```{r}
#| eval: true

DBI::dbExecute(
  conn      = pg,
  statement = "
  ALTER TABLE stems_temp.plots_plants
    ALTER COLUMN survey_date TYPE DATE USING survey_date::date,
    ADD COLUMN shrub_id INTEGER,
    ADD COLUMN stem_id_new INTEGER,
    ADD COLUMN stem_id_old INTEGER
  ;
  "
)

```


# add stem_id generated from update of old stems to plots_plants (a later step)

```{r}
#| eval: true

update_plots_plants <- "
  UPDATE stems_temp.plots_plants
  SET shrub_id = urbancndep.shrubs.id
  FROM urbancndep.shrubs
  WHERE
    urbancndep.shrubs.plot_id = plots_plants.plot_id AND
    urbancndep.shrubs.code = plots_plants.plant_id
  ;
  "

safe_transaction(
  action     = update_plots_plants,
  ref_object = plots_plants
)

```


# update existing pg.stems with this survey's post date & notes

Be certain to set month & year to the correct pre date, which should be the most
recent set (may include multiple months).

Get the month and date of the most recent previous stems survey to pass to the
update query. Note that this query is specific to when the last set of stems
were measured in a single month (e.g., 5 or 10). If stem surveys spanned
multiple months, which is rare but has happened, you either need to update the
query or pass those values manually.

```{r}
#| eval: true

recent_pre_date <- DBI::dbGetQuery(
  conn      = pg,
  statement = "
  SELECT
    DISTINCT EXTRACT(MONTH FROM pre_date) AS recentmos,
    EXTRACT(YEAR FROM pre_date) AS recentyear
  FROM
    urbancndep.stems
  WHERE
    EXTRACT(YEAR FROM pre_date) = (SELECT MAX(EXTRACT(YEAR FROM pre_date)) FROM urbancndep.stems)
  ;
  "
)

recentmos <- sort(unique(as.integer(recent_pre_date$recentmos)))
recentmos_gt8 <- recentmos[recentmos > 8]
recentmos_lt8 <- recentmos[recentmos < 8]

# Rule: prefer all months > 8 and ignore < 8 whenever any > 8 exist.
# Otherwise, use all months < 8 (only when there are no months > 8).
if (length(recentmos_gt8) > 0) {
  pre_date_month <- recentmos_gt8
  month_rule <- "using recentmos > 8"
} else if (length(recentmos_lt8) > 0) {
  pre_date_month <- recentmos_lt8
  month_rule <- "using recentmos < 8 (no recentmos > 8 found)"
} else {
  stop("No valid values found in recentmos for pre_date_month selection.")
}

pre_date_year <- max(recent_pre_date$recentyear)

message("month selection rule: ", month_rule)
message("pre_date_month: ", paste(pre_date_month, collapse = ", "))
message("pre_date_year: ", pre_date_year)

```

to address manually, change the last two lines of update_post_query to:
EXTRACT(MONTH from urbancndep.stems.pre_date) = __SET_PREVIOUS_PERIOD_MONTH__ AND
EXTRACT(YEAR from urbancndep.stems.pre_date) = __SET_PREVIOUS_PERIOD_YEAR;") # set appropriately !

```{r}
#| eval: true

update_post_query <- glue::glue_sql("
  UPDATE urbancndep.stems
  SET
    post_date = plots_plants.survey_date,
    post_note = plots_plants.post_note
  FROM stems_temp.plots_plants
  WHERE
    plots_plants.shrub_id = urbancndep.stems.shrub_id AND
    plots_plants.direction = urbancndep.stems.direction AND
    EXTRACT(MONTH from urbancndep.stems.pre_date) IN ({ pre_date_month* }) AND
    EXTRACT(YEAR from urbancndep.stems.pre_date) = { pre_date_year }
  ;
  ",
  .con = DBI::ANSI()
)

safe_transaction(
  action     = update_post_query,
  ref_object = plots_plants
)

```

# insert pre data into pg.stems

```{r}
#| eval: true

insert_pre <- "
  INSERT INTO urbancndep.stems
  (
    shrub_id,
    direction,
    pre_date
  )
  (
    SELECT
      shrub_id,
      direction,
      survey_date
    FROM stems_temp.plots_plants
  )
  ;
  "

safe_transaction(
  action     = insert_pre,
  ref_object = plots_plants
)

```


# add stem_id generated from insert of new stems to plots_plants

note the use of `urbancndep.stems.pre_date` in this query

```{r}
#| eval: true

add_stem_id_new <- "
  UPDATE stems_temp.plots_plants
  SET stem_id_new = urbancndep.stems.id
  FROM urbancndep.stems
  WHERE
    urbancndep.stems.shrub_id = plots_plants.shrub_id AND
    urbancndep.stems.direction = plots_plants.direction AND
    urbancndep.stems.pre_date = plots_plants.survey_date
  ;
  "

safe_transaction(
  action     = add_stem_id_new,
  ref_object = plots_plants
)

```

# add stem_id generated from update of old stems to plots_plants

```{r}
#| eval: true

add_stem_id_old <- "
  UPDATE stems_temp.plots_plants
  SET stem_id_old = urbancndep.stems.id
  FROM urbancndep.stems
  WHERE
    urbancndep.stems.shrub_id = plots_plants.shrub_id AND
    urbancndep.stems.direction = plots_plants.direction AND
    urbancndep.stems.post_date = plots_plants.survey_date
  ;
  "

safe_transaction(
  action     = add_stem_id_old,
  ref_object = plots_plants
)

```


# prep for and add stem_id generated from update of stems to old;

```{r}
#| eval: true

DBI::dbExecute(
  conn      = pg,
  statement = "
  ALTER TABLE stems_temp.old
    ADD COLUMN stem_id INTEGER,
    ADD COLUMN postmeas boolean
  ;
  "
)

```


```{r}
#| eval: true

update_stems_old <- "
  UPDATE stems_temp.old
  SET stem_id = plots_plants.stem_id_old
  FROM stems_temp.plots_plants
  WHERE
    plots_plants.plant_id = old.plant_id AND
    plots_plants.plot_id = old.plot_id AND
    SUBSTRING(plots_plants.direction, 1, 1) = old.old_direction
  ;
  "

safe_transaction(
  action     = update_stems_old,
  ref_object = old
)

```

# add boolean TRUE to denote post measurement

```{r}
#| eval: true

DBI::dbExecute(
  conn      = pg,
  statement = "
  UPDATE stems_temp.old
  SET postmeas=TRUE
  ;
  "
)
```


# insert old lengths into pg.stem_lengths

```{r}
#| eval: true

insert_old <- "
  INSERT INTO urbancndep.stem_lengths
  (
    stem_id,
    length_in_mm,
    post_measurement
  )
  (
    SELECT
      stem_id,
      old_length::DOUBLE PRECISION,
      postmeas
    FROM stems_temp.old
    WHERE old_length IS NOT NULL
  )
  ;
  "

safe_transaction(
  action     = insert_old,
  ref_object = old
)

```

# prep for and add stem_id generated from insert of stems to new;
# prep boolean FALSE to denote post measurement (later step)

```{r}
#| eval: true

DBI::dbExecute(
  conn      = pg,
  statement = "
  ALTER TABLE stems_temp.new
    ADD COLUMN stem_id integer,
    ADD COLUMN postmeas boolean
  ;
  "
)

```


```{r}
#| eval: true

update_stems_new <- "
  UPDATE stems_temp.new
  SET stem_id = plots_plants.stem_id_new
  FROM stems_temp.plots_plants
  WHERE
    plots_plants.plant_id = new.plant_id AND
    plots_plants.plot_id = new.plot_id AND
    SUBSTRING(plots_plants.direction, 1, 1) = new.new_direction
  ;
  "

safe_transaction(
  action     = update_stems_new,
  ref_object = new
)

```


```{r}
#| eval: true

DBI::dbExecute(
  conn      = pg,
  statement = "
  UPDATE stems_temp.new
  SET postmeas = FALSE
  ;
  "
)

```


# insert new lengths into pg.stem_lengths

```{r}
#| eval: true

insert_stems_new <- "
  INSERT INTO urbancndep.stem_lengths
  (
    stem_id,
    length_in_mm,
    post_measurement
  )
  (
    SELECT
      stem_id,
      new_length::DOUBLE PRECISION,
      postmeas
    FROM stems_temp.new
    WHERE new_length is not null
    )
  ;
  "

safe_transaction(
  action     = insert_stems_new,
  ref_object = new
)

```


# insert any notes about plants into stem_comment using plant-level keys
# (shrub_id, survey_date). Values are deduplicated per key when the same note
# appears in multiple directional rows.

```{r}
#| eval: true

DBI::dbWithTransaction(
  conn = pg,
  {
    DBI::dbExecute(
      conn      = pg,
      statement = "
      INSERT INTO urbancndep.stem_comment
      (
        shrub_id,
        survey_date,
        comment
      )
      (
      WITH stems_temp_note_rows AS (
        SELECT
          stems_temp.plots_plants.shrub_id,
          stems_temp.plots_plants.survey_date::date AS survey_date,
          NULLIF(BTRIM(stems_temp.plots_plants.note_about_plant), '') AS comment
        FROM stems_temp.plots_plants
        WHERE stems_temp.plots_plants.shrub_id IS NOT NULL
          AND stems_temp.plots_plants.survey_date IS NOT NULL
          AND NULLIF(BTRIM(stems_temp.plots_plants.note_about_plant), '') IS NOT NULL
      )
      SELECT
        stems_temp_note_rows.shrub_id,
        stems_temp_note_rows.survey_date,
        STRING_AGG(
          DISTINCT stems_temp_note_rows.comment,
          '; ' ORDER BY stems_temp_note_rows.comment
        ) AS comment
      FROM stems_temp_note_rows
      GROUP BY stems_temp_note_rows.shrub_id, stems_temp_note_rows.survey_date
      )
      ;
      "
    )
  }
)
```


# identify plot notes that will be rejected before insert

```{r}
#| eval: true

DBI::dbGetQuery(
  conn = pg,
  statement = "
  SELECT
    COUNT(*) AS rejected_plot_note_rows
  FROM stems_temp.plot_notes
  WHERE survey_date IS NULL
     OR NULLIF(BTRIM(plot_notes), '') IS NULL
  ;
  "
)

```


# add plot notes (new as of 2018-10-22)

```{r}
#| eval: true

DBI::dbWithTransaction(
  conn = pg,
  {
    DBI::dbExecute(
      conn      = pg,
      statement = "
      INSERT INTO urbancndep.stem_plot_notes
      (
        plot_id,
        survey_date,
        plot_notes
      )
      (
      SELECT
        plot_id,
        survey_date::date,
        NULLIF(BTRIM(plot_notes), '') AS plot_notes
      FROM stems_temp.plot_notes
      WHERE survey_date IS NOT NULL
        AND NULLIF(BTRIM(plot_notes), '') IS NOT NULL
      )
      ;
      "
    )
  }
)
```
          

# address missing NULL values ---------------------------------------------

WHILE THE WORKFLOW TO ADDRESS MISSING NEW VALUES WAS IMPROVED WITH THE FALL 2022
WORKFLOW, THERE WERE NOT ANY NULL VALUES ASSOCIATED WITH THAT COLLECTION SO
ADDRESS A WORKFLOW FOR NULL NEW VALUES IN A FUTURE COLLECTION.
- still the case through the 2023 data.
- still the case through the 2024 (spring) data.
- still the case through the 2024 (fall) data.

Identify the details of any NULL value new stems using new_null(). The trick
here is that this should only be an issue if there is a new stem with a single
null value and no other values. If that is the case, you may have to treat that
stem like a missing new value. However, if there are two new values for a single
direction, then one can be NULL but the other one will be entered into the
database and thus avoid having a missing direction. So, for example, in May
2017, there was a NULL value at WTM-71-L1-E but there was another measurement
for E so we could simple ignore that NULL value

```{r}
#| eval: true

new_null <- function() {

  newLengths <- new

  dirSym <- c("N", "S", "W", "E")
  dirSym <- data.frame(dirSym, stringsAsFactors = FALSE)

  nullvalues <- newLengths |>
    dplyr::filter(is.na(new_length))

  if (nrow(nullvalues) > 0) {

    aframe <- data.frame(stringsAsFactors = FALSE)

    for (i in 1:nrow(nullvalues)) {
      plot <- nullvalues[i, ]$plot_id
      plant <- nullvalues[i, ]$plant_id

      aframe <- dplyr::bind_rows(
        aframe,
        newLengths |>
          dplyr::filter(plot_id == plot, plant_id == plant) |>
          dplyr::right_join(dirSym, by = c("new_direction" = "dirSym")) |>
          dplyr::mutate(plot_id = replace(plot_id, is.na(new_length), plot)) |>
          dplyr::mutate(plant_id = replace(plant_id, is.na(new_length), plant))
      )
    }

    output <- aframe |>
      dplyr::arrange(plot_id, plant_id, new_direction)

  } else {

    output <- NULL

  }

  return(output)

}

new_null()
```

# address missing NEW values ----------------------------------------------

identify and document missing new stems data

Identify the details of any missing new stems using \code{identify_new_missing}.
We need to document cases where there are not any new-stem measurements for a
particular `plot*plant*direction`. We can then pass these details to
\code{annotate_new_missing} to: (1) document the NULL value for that direction,
and (2) add a `missing value` note to `urbancndep.stems.pre_note`.

```{r}
#| eval: true

new_stems_missing <- identify_new_missing()

purrr::pwalk(
  .l = new_stems_missing,
  .f = ~ annotate_new_missing(plot_id = ..4, plant_id = ..1, direction = ..2)
)
```


# post-processing error checking -----------------------------------------------

```{r}
#| eval: true

years <- c(
  lubridate::year(Sys.Date()) - 2,
  lubridate::year(Sys.Date()) - 1,
  lubridate::year(Sys.Date())
)

query_recent <- glue::glue_sql("
  SELECT
    s.code AS site_code,
    p.id AS plot_id,
    t.code AS treatment_code,
    sp.scientific_name,
    sh.code AS shrub_code,
    sh.note AS shrub_note,
    st.id,
    st.direction,
    st.pre_date,
    st.pre_note,
    st.post_date,
    st.post_note,
    sl.post_measurement,
    sl.length_in_mm as stem_length,
    sc.comment
  FROM urbancndep.stems st
  JOIN urbancndep.shrubs sh ON sh.id = st.shrub_id
  JOIN urbancndep.plots p ON sh.plot_id = p.id
  JOIN urbancndep.sites s ON p.site_id = s.id
  JOIN urbancndep.treatments t ON p.treatment_id = t.id
  JOIN urbancndep.shrub_species sp ON sh.shrub_species_id = sp.id
  LEFT JOIN urbancndep.stem_lengths sl ON st.id = sl.stem_id
  LEFT JOIN urbancndep.stem_comment sc ON (sc.stem_id = st.id AND sl.post_measurement = sc.post_measurement)
  WHERE
    EXTRACT (YEAR FROM st.pre_date) IN ({ years* })
  ORDER BY
    sl.post_measurement,
    p.id,
    st.pre_date,
    sh.code,
    st.direction
  ;
  ",
  .con = DBI::ANSI()
)

recent_data <- DBI::dbGetQuery(
  conn      = pg,
  statement = query_recent
)
```


# shrub dimensions -------------------------------------------------------------

```{r}
#| eval: true

DBI::dbExecute(
  conn      = pg,
  statement = "
  ALTER TABLE stems_temp.shrub_dimensions
  ADD COLUMN shrub_id INTEGER
  ;
  "
)
```


```{r}
#| eval: true

DBI::dbWithTransaction(
  conn = pg,
  {
    DBI::dbExecute(
      conn      = pg,
      statement = "
      UPDATE stems_temp.shrub_dimensions
      SET shrub_id = urbancndep.shrubs.id
      FROM urbancndep.shrubs
      WHERE
        urbancndep.shrubs.plot_id = shrub_dimensions.plot_id AND
        urbancndep.shrubs.code = shrub_dimensions.plant_id
      ;
      "
    )
    DBI::dbExecute(
      conn      = pg,
      statement = "
      INSERT INTO urbancndep.shrub_measurements
      (
        plot_id,
        survey_date,
        plant,
        canopy_extent_n_s,
        canopy_extent_e_w,
        height,
        notes,
        shrub_id
      )
      (
      SELECT
        plot_id,
        survey_date::DATE,
        'Larrea tridentata',
        n_s,
        e_w,
        height_of_plant,
        note_about_plant,
        shrub_id
      FROM stems_temp.shrub_dimensions
      )
      ;
      "
    )
  }
)
```

# check shrub dimensions data

```{r}
#| eval: true

query_recent_measurements <- glue::glue_sql("
  SELECT
    sites.code AS site_code,
    plots.id AS plot_id,
    shrub_measurements.survey_date,
    shrub_measurements.canopy_extent_n_s AS n_s,
    shrub_measurements.canopy_extent_e_w AS e_w,
    shrub_measurements.height,
    shrubs.code AS shrub_code,
    shrub_measurements.notes
  FROM urbancndep.shrub_measurements
  LEFT JOIN urbancndep.shrubs ON (shrubs.id = shrub_measurements.shrub_id)
  LEFT JOIN urbancndep.plots ON (shrubs.plot_id = plots.id)
  LEFT JOIN urbancndep.sites ON (plots.site_id = sites.id)
  WHERE EXTRACT (YEAR FROM shrub_measurements.survey_date) IN ({ years* })
  ORDER BY
    plots.id,
    shrubs.code
  ;
  ",
  .con = DBI::ANSI()
)

recent_measurements <- DBI::dbGetQuery(
  conn      = pg,
  statement = query_recent_measurements
)
```

# CLEAN UP ---------------------------------------------------------------------

```{r}
#| eval: true

DBI::dbExecute(
  conn      = pg,
  statement = "DROP SCHEMA IF EXISTS stems_temp CASCADE ;"
)
```
