---
title: "populate_database"
format: gfm
execute:
  enabled: true
---

# README

Workflow to upload DesFert stem-length data to the `urbancndep` database.

The KoBo transformation and validation run before staging. Production database
writes then run in one transaction: any failed mapping, affected-row check, or
SQL statement rolls the entire collection upload back.

# Prepare the R session

The connection settings must assign `pg` from `RPostgres::Postgres()`. The
workflow rejects legacy `RPostgreSQL::PostgreSQL()` connections before reading
or writing database objects because their handling of schema-qualified
identifiers differs.

```{r}
#| label: database connection
#| eval: true

source("~/Documents/localSettings/pg_local.R")
source("helper_check_database_connection.R")

check_rpostgres_connection(
  connection = pg,
  connection_name = "upload connection"
)

```

```{r}
#| label: source kobo workflow
#| eval: true

source("kobo_workflow.R")

```

```{r}
#| label: load database helpers
#| eval: true

source("helper_load_table.R")
source("helper_safe_transaction.R")
source("helper_upload_stems_data.R")

```

# Confirm the database target

```{r}
#| label: database target
#| eval: true

database_target <- DBI::dbGetQuery(
  conn = pg,
  statement = "
    SELECT
      current_database() AS database_name,
      current_user AS database_user,
      inet_server_addr()::text AS server_address
    ;
  "
)

base::print(database_target)

if (!base::identical(database_target$database_name[[1]], "caplter")) {
  base::warning(
    "database target is not caplter",
    call. = FALSE
  )
}

```

# Build and load staging tables

The staging schema contains only derived data for the current upload. Production
tables are not changed during this phase.

```{r}
#| label: build staging schema
#| eval: true

DBI::dbExecute(
  conn = pg,
  statement = "DROP SCHEMA IF EXISTS stems_temp CASCADE;"
)
DBI::dbExecute(
  conn = pg,
  statement = "CREATE SCHEMA stems_temp;"
)

```

```{r}
#| label: load staging tables
#| eval: true

staging_data <- base::list(
  new = new,
  old = old,
  plots_plants = plots_plants,
  plot_notes = plot_notes,
  shrub_dimensions = shrub_dimensions,
  new_stems_missing = new_stems_missing
)

staging_results <- purrr::imap_dfr(
  .x = staging_data,
  .f = ~ helper_load_table(
    connection = pg,
    schema_name = "stems_temp",
    table_name = .y,
    table_data = .x
  )
)

base::print(staging_results)

```

# Select the preceding collection

There are two collections per year. Prefer fall months when the most recent
database year contains fall records; otherwise use its spring months.

```{r}
#| label: select preceding collection
#| eval: true

recent_pre_date <- DBI::dbGetQuery(
  conn = pg,
  statement = "
    SELECT DISTINCT
      EXTRACT(MONTH FROM stems.pre_date) AS recent_month,
      EXTRACT(YEAR FROM stems.pre_date) AS recent_year
    FROM urbancndep.stems AS stems
    WHERE EXTRACT(YEAR FROM stems.pre_date) = (
      SELECT MAX(EXTRACT(YEAR FROM prior_stems.pre_date))
      FROM urbancndep.stems AS prior_stems
    )
    ;
  "
)

recent_months <- base::sort(
  base::unique(base::as.integer(recent_pre_date$recent_month))
)
fall_months <- recent_months[recent_months > 8]
spring_months <- recent_months[recent_months < 8]

if (base::length(fall_months) > 0) {
  pre_date_month <- fall_months
  month_rule <- "fall months from the most recent database year"
} else if (base::length(spring_months) > 0) {
  pre_date_month <- spring_months
  month_rule <- "spring months from the most recent database year"
} else {
  base::stop(
    "no valid preceding collection months were found",
    call. = FALSE
  )
}

pre_date_year <- base::max(recent_pre_date$recent_year)

base::message("month selection rule: ", month_rule)
base::message(
  "pre_date_month: ",
  base::paste(pre_date_month, collapse = ", ")
)
base::message("pre_date_year: ", pre_date_year)

```

# Upload the collection atomically

The source checksum prevents a successfully committed export from being
uploaded twice. Database-generated stem IDs are mapped and consumed inside the
same transaction; they do not require intermediate commits.

```{r}
#| label: upload stems data
#| eval: true

source_md5 <- base::unname(tools::md5sum(path))

upload_stats <- upload_stems_data(
  db_connection = pg,
  pre_date_month = pre_date_month,
  pre_date_year = pre_date_year,
  source_file = base::basename(path),
  source_md5 = source_md5
)

base::print(upload_stats)

```

# Post-upload verification

## Recent stem data

Plant comments use `(shrub_id, survey_date)`. The applicable survey date is
selected from the stem's pre/post event according to the length record.

```{r}
#| label: verify recent stem data
#| eval: true

years <- c(
  lubridate::year(base::Sys.Date()) - 2,
  lubridate::year(base::Sys.Date()) - 1,
  lubridate::year(base::Sys.Date())
)

query_recent <- glue::glue_sql(
  "
  SELECT
    sites.code AS site_code,
    plots.id AS plot_id,
    treatments.code AS treatment_code,
    shrub_species.scientific_name,
    shrubs.code AS shrub_code,
    shrubs.note AS shrub_note,
    stems.id AS stem_id,
    stems.direction,
    stems.pre_date,
    stems.pre_note,
    stems.post_date,
    stems.post_note,
    stem_lengths.post_measurement,
    stem_lengths.length_in_mm AS stem_length,
    stem_comment.comment
  FROM urbancndep.stems AS stems
  JOIN urbancndep.shrubs AS shrubs
    ON shrubs.id = stems.shrub_id
  JOIN urbancndep.plots AS plots
    ON plots.id = shrubs.plot_id
  JOIN urbancndep.sites AS sites
    ON sites.id = plots.site_id
  JOIN urbancndep.treatments AS treatments
    ON treatments.id = plots.treatment_id
  JOIN urbancndep.shrub_species AS shrub_species
    ON shrub_species.id = shrubs.shrub_species_id
  LEFT JOIN urbancndep.stem_lengths AS stem_lengths
    ON stem_lengths.stem_id = stems.id
  LEFT JOIN urbancndep.stem_comment AS stem_comment
    ON stem_comment.shrub_id = stems.shrub_id
   AND stem_comment.survey_date = CASE
     WHEN stem_lengths.post_measurement THEN stems.post_date
     ELSE stems.pre_date
   END
  WHERE EXTRACT(YEAR FROM stems.pre_date) IN ({years*})
  ORDER BY
    stem_lengths.post_measurement,
    plots.id,
    stems.pre_date,
    shrubs.code,
    stems.direction
  ;
  ",
  .con = pg
)

recent_data <- DBI::dbGetQuery(
  conn = pg,
  statement = query_recent
)

data.table::fwrite(
  x = recent_data,
  file = "/tmp/recent_data.csv",
  dateTimeAs = "write.csv"
)

```

## Recent shrub measurements

```{r}
#| label: verify recent shrub measurements
#| eval: true

query_recent_measurements <- glue::glue_sql(
  "
  SELECT
    sites.code AS site_code,
    plots.id AS plot_id,
    shrub_measurements.survey_date,
    shrub_measurements.canopy_extent_n_s AS n_s,
    shrub_measurements.canopy_extent_e_w AS e_w,
    shrub_measurements.height,
    shrubs.code AS shrub_code,
    shrub_measurements.notes
  FROM urbancndep.shrub_measurements AS shrub_measurements
  LEFT JOIN urbancndep.shrubs AS shrubs
    ON shrubs.id = shrub_measurements.shrub_id
  LEFT JOIN urbancndep.plots AS plots
    ON plots.id = shrubs.plot_id
  LEFT JOIN urbancndep.sites AS sites
    ON sites.id = plots.site_id
  WHERE EXTRACT(YEAR FROM shrub_measurements.survey_date) IN ({years*})
  ORDER BY
    plots.id,
    shrubs.code
  ;
  ",
  .con = pg
)

recent_measurements <- DBI::dbGetQuery(
  conn = pg,
  statement = query_recent_measurements
)

data.table::fwrite(
  x = recent_measurements,
  file = "/tmp/recent_measurements.csv",
  dateTimeAs = "write.csv"
)

```

# Clean up staging

This step runs only after the atomic upload and verification chunks succeed.

```{r}
#| label: drop staging schema
#| eval: true

DBI::dbExecute(
  conn = pg,
  statement = "DROP SCHEMA IF EXISTS stems_temp CASCADE;"
)

```
